What's hot and what's not:\\Windowing developer topic analysis

### 2009-04-10 lda-paper.tex
More edits and changes cleaned up descriptions

We mirrored the repositories and their revisions using software such
as rsync, CVSSuck, \texttt{softChange} and bt2csv.
\texttt{softChange} provided a general schema for storing
revisions. CVSSuck and rsync mirrored CVS repositories while bt2csv
mirrored web accessible bitkeeper repositories.

@@

The document we are analyzing are the changelog comments i.e., the comments that are added  when a revision to the source code is committed.
Documents are converted into word count distributions. These
distributions, which are then normalized by by the size of the
message. After all messages are processed the distributions are
extended to include all words in all distributions (each distribution
is extended with word bins of count 0).

@@

Given all messages, we group the messages into windows. We could use
overlapping windows, but in this paper we use non-overlapping windows
of a set time unit because it simplifies analysis.  Windowing by time
allows for many levels of granularity. We usually used one month as
the length of our time windows, while we could use different
granularities for the study we think that a month is a sizable unit of
development, 4 weeks of development, which we consider to be fine
enough to show meaningful shifts but coarse enough not to show too
much detail.

@@

Once we have our data windowed, we apply our Topic Analysis tool to
each window and extract the top $N$ topic clusters, we used $20$
topics in our case study. Our Topic Analysis tool is an implementation
of LDA, but we could have used other similar tools like LSI, but
previous showed more promising topic analysis with LDA.

This is a very slow step as LDA takes a long time to run, and takes
even longer since we run it once per window. We note that this step is
a slow one, as executing even one instance of LDA involves significant
computation, and we perform LDA once per window.

@@

We then abstract these topics by comparing them to each other.  Each
topic is compared to each other topic in the system, given a certain
threshold of top 10 similarity, such as 8/10 words. Then given this topic
similarity matrix, we find the transitive closure of similar topics;
that is, we fill flood along the similarity arcs until we have
partitioned the topics into clusters of similar topics. Figure
\ref{fig:closure} provides an example of clustering topics by the
topic similarity. These clusters of topics are called trends. A trend
is probably interesting and valuable if there is more than one topic in
the trend.

@@

We have devised several ways to visualize these topics and
trends.  For all of these plots if we find trends that have continuous
segments, then we plot those segments as joined across time.  We have a
compressed stacked trend view, shown in Figure \ref{fig:topicsmear}
and Figure \ref{fig:zoomedsmear}, 
which displays trends as they occur
on the time-based x-axis (placement along the y-axis is arbitrary).
 Another plot we have is the trend-time line plot, shown in
Figure \ref{fig:trendtimeline}, each trend gets it own distinct y-value ,
while the x-axis is time; these topics are then each plotted on their
own line across time as they occur. Our final plot is the histogram
plot, shown in Figure \ref{fig:histogram} where we plot each trend on
its own line but stack up the segments of the trend, much like a plot
of counts or a histogram.

@@

The second visualization is the trend-time line plot (Figure
\ref{fig:trendtimeline}), it displays repeating trends more clearly by
dedicating a horizontal line for trend segments dedicated to one
trend. This means that if a trend contains discontinuous segments then the
segments appear on the same line. 
However, the least common trends need to be pruned away or the view will be cluttered with a lot of noise.

The third visualization is the histogram plot (Figure
\ref{fig:histogram}) which superficially resembles the time-trend
plot.  However, in this view the trends are plotted together by
stacking to their row so time information is lost, and the trends are
ordered by the number of topics in the trend.  The histogram plot
shows the count of instances of a trend and thus indicates which
trends occur the most. Unfortunately due to the large number of topics
given allotted space it is often best to crop off the trends with only
one topic, otherwise the tail is long.

%XXX Describe something useful like a trend
\begin{figure*}
  \centering
  \includegraphics[width=1.0\textwidth]{fixed-time-smear-plot-scaled}
  \caption{Compact Trend Plot Topics per Month for MaxDB 7.500. The
    x-axis is time in months, the Y axis is used to stack topics
    occurring at the same time. Trends that are continuous are
    plotted as continuous blocks. The top 10 words in the topics are
    joined and embedded in box representing the topics.}
  \label{fig:topicsmear}
\end{figure*}

\begin{figure}
  \centering
  \includegraphics[width=0.45\textwidth]{fixed-time-smear-plot-cropped}
  \caption{Zoomed in Slice Compact Trend Plot of Topics per Month for MaxDB 7.500. The topic text is visible in each topic box. Trends are plotted across time continuously.}
  \label{fig:zoomedsmear}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.45\textwidth]{class-smear-plot-crop-scaled}
  \caption{Trend-time Line: Trends plotted over time of MaxDB 7.500. Time in months are plotted along the x-Axis, each row on the y-axis is associated with a trend ranked by size in descending order}         
  \label{fig:trendtimeline}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.45\textwidth]{histogram-cropped-scaled}
  \caption{Top part of trend histogram, ordered by topic occurrence for MaxDB 7.500. X axis determines the number of continuous months of a trend. Trends are ranked  by the number of topics that a trend contains in descending order.}       
  \label{fig:histogram}
\end{figure}

@@

When we examined PostgreSQL, we did not find many trends with two or more topics; however, we did find one large trend which stretched across most of its existence.
 Much of the large trend consisted of stop words but it did
include topics which had non-stop words such as \emph{patch}, \emph{fix}, \emph{update}.

The second largest topic was interesting because it concerned changes from
1998, from developers using Central Europe Time (CEST), where the two
notable tokens were 1998 and CEST. These changes appear to be patches
submitted by one particular developer who lived in that timezone.

The third largest trend we found appeared to concerned with copying
and moving files between branches as it referenced the old branch.
branches.  We inferred this trend based on the fact that the old
version identifier was referenced, and we know this this is common
practice when performing a branch merge, since CVS cannot record such
information directly.

Once stop words were removed the some of the purposes became more
clear. The large trend disappeared, the largest trend became bug fixes
submitted by a developer named Dal Zotto which is a short trend.

@@

We tracked Firebird from August 2000 to January 2006.
Our first attempt with Firebird resulted in a set of trends that were
dominated by one large trend containing tokens that were mostly stop
words such as references to one author
\texttt{carlosga05} and the words: \emph{the, to, a, for, of, is, it}.

The second largest trend related to branching and possibly merging; it
contained language such as ``was added on branch'' and contained
references to tags and branches.

The third largest trend contained references to the changelog, the
build system, architecture and platform support. It specifically
mentioned tokens such as \emph{Solaris}, \emph{port} and \emph{x86}.

Other trends included topics regarding AIX PPC compilation, updating
the build process, internationalization and UTF8, Darwin build support
and bug fixing.

Topics that were not trends but appeared to be  interesting were mostly external bug
fixes submitted to the project.
 In these cases, the developers would express thanks in their changelog comments, such as ``Thanks, Bill Lam''.  Other easily discernible topics included 
 tokens and topics such as: \emph{compiler workarounds}, \emph{nightly updates}, \emph{packets} and \emph{MSVC++}.

@@

The plots we produced of MaxDB 7.500 were unlike those of the other
systems, as there was a period where no real development occurred and
thus there were no topics or trends whatsoever. We have two periods to
evaluate for MaxDB 7.500, the first period from June 2004 to January 2005,
and the second period from June 2005 to June 2006.

The largest common trend has references to build system files like
SYSDD.punix and MONITOR.punix.  This trend is partially displayed at
the top of the zoomed in compressed view plot (Figure
\ref{fig:zoomedsmear}) and at the top of the histogram plot (Figure
\ref{fig:histogram}).  Other tokens mentioned are Sutcheck v1.09, the
prefix SUT stands for Storable Unit Type. Sutcheck would also automate
check-ins using a Perforce SCS tool which was exported to CVS.

@@

The third most common trend, seen on Figure \ref{fig:trendtimeline},
seemed to include tokens related to operating system support, such as
Linux and Windows, as well as architecture support, AMD64 and
Opteron. The word \emph{problem} was common among all of these
trends. This trend seemed related to the smaller fourth largest trend
which occurred nearby which referenced tokens \emph{AMD64} and
\emph{Windows}, this shows that topics can overlap but still not match
via our similarity measure.

%XXX Edit
% Was reading lda.report
Bug tracker URLs dominated unique topics during some months. For
instance in the last month of MaxDB 7.500 development every topic
contained 1 unique Bug tracker URL, this pattern did not occur in the
previou month. We investigated the actual revisions and we did find
there was difference in behaviour between and that the last month of
development commits were mostly referencing the bug tracker. If the
topics of one month were 20 or more unique bug tickets being
addressed, the global LDA analysis would probably miss this, yet these
bug tickets were the focus of development for that month and not
necessarily globally relevant.

@@

We noticed that PHP was in a third of the commits that Perl was in
(not the same commits) in as many commits as Perl yet PHP did not
appear in the global static topic analysis topics while Perl did. Our
local topic analysis mentioned PHP in 5 different topics, yet
mentioned Perl in 4 different topics. Perhaps the Perl mentions were
not as independent topics as the PHP topics were.

We noticed that Perforce and Changelist were in just about every
topic, so we added them to the stop words list. Longer trends were
shortened and in some cases the total number of topics found per month
was less. Evaluating different similarity thresholds showed that by
removing common words one reduces the general similarity of topics,
that said the larger topics still existed. Thus if you add more
relevant stop words one should tune your topic similarity parameters
to handle this change.

@@

If we approach the difference via common tokens we can see that common
token tend to dominate both results, at least in the case of MaxDB
7.500 our local topic approach netted us topics which contained these
relatively important and common words: \emph{UNICODE},
\emph{DEC/OSF1}, \emph{ODBC}, \emph{crash}, \emph{SQLCLI},
\emph{SYSDD.cpnix}, \emph{backup}, \emph{select}, \emph{make},
\emph{memory}, \emph{view}, and finally \emph{debug}. None of these
tokens were part of the static topic analysis topics, but there were
part of 566 commits (6\% of the entire system) to MaxDB 7.500 which
contained these. These tokens were part of 87 out of 500 of our
locally generated topics (ignoring similar topics).

@@

Each project seemed to have their own kind of stop words, we did not
remove them, but perhaps dropping some of these stop words would aid
the clarity of such topic cluster; alternatively it might cause our
topic similarity plots to be fundamentally different.

@@

We demonstrated the application of windowed topic analysis, topic
analysis applied to commit messages during periods of
development. This approach differed from previous work, as previous
work would apply topic analysis globally and then analyze related
documents. Instead we show that many topics that exist locally are
relevant and interesting yet would often not be included in global
topic analysis. We handle the case of recurring topics by using a
topic similarity measure which looks for topics which recur and mutate
throughout the development of the software project.

The value of windowed topic analysis was demonstrated on a case study
of three database systems with a particular focus on MaxDB 7.500. We
showed that static topic analysis missed important database concepts
such as ODBC.

@@

We presented multiple visualizations that focused on different aspects
of the data: temporality of trends, trend size, and a condensed
overall trend view. The condensed trend view shows much more
information than the views that global analysis could show, it
indicates how focused a period is by the total number of topics as
well shows topics by similarity so one can track trends across time.
Our histogram trend-view immediately hi-lights and measures the size of
trends while our time-line trend view shows how a topic reoccurs over
time. These visualizations help us get an over feeling about the
common and unique topics that developers focus on. If implemented
properly a user can easily zoom and query a topic.

>>>>

We mirrored the repositories and their revisions using software such
as rsync, CVSSuck, \texttt{softChange} and bt2csv.
\texttt{softChange} provided a general schema for storing
revisions. CVSSuck and rsync mirrored CVS repositories while bt2csv
mirrored web accessible BitKeeper repositories.

@@

The documents we are analyzing are the commit log comments i.e., the
comments that are added when revisions to the project are committed.
Documents are converted into word count distributions. These
distributions are then normalized by the size of the message. After
all messages are processed the distributions are extended to include
all words within all of the  distributions.

@@

Given all messages, we group the messages into windows. We could use
overlapping windows, but in this paper we use non-overlapping windows
of a set time unit because it simplifies analysis.  Windowing by time
allows for many levels of granularity. We usually used one month as
the length of our time windows, while we could use different
granularities for the study we think that a month is a sizable unit of
development, which we consider to be fine enough to show meaningful
shifts in focus, but coarse enough not to show too much detail.

@@

Once we have our data windowed, we apply our Topic Analysis tool to
each window and extract the top $N$ topic clusters, we used $20$
topics in our case studies. Our Topic Analysis tool is an
implementation of LDA, but we could have used other similar tools like
LSI, but previous work showed more promising topic analysis results
with LDA.  We note that this step is a slow one, as executing even one
instance of LDA involves significant computation, and we perform LDA
once per window.

@@

We then abstract these topics by comparing them to each other.  Each
topic is compared to each other topic in the system, given a certain
threshold of top 10 similarity, such as 8 out of 10 words. Then given
this topic similarity matrix, we find the transitive closure of
similar topics; that is, if we model topics are nodes and arcs as
similarity, we fill flood along the similarity arcs until we have
partitioned the topics into clusters of similar topics. Figure
\ref{fig:closure} provides an example of clustering topics by the
topic similarity. These clusters of topics are called trends. A trend
is probably interesting and valuable if there is more than one topic
in the trend.

@@

We have devised several ways to visualize these topics and trends.
For all of these plots if we find trends that have continuous
segments, then we plot those segments as joined across time.  We have
a compressed stacked trend view, shown in Figure \ref{fig:topicsmear}
and Figure \ref{fig:zoomedsmear}, which displays trends as they occur
on the time-based x-axis (placement along the y-axis is arbitrary).
Another plot we have is the trend-time line plot, shown in Figure
\ref{fig:trendtimeline}, each trend gets it own distinct y-value ,
while the x-axis is time; these topics are then each plotted on their
own line across time as they occur. Our final plot is the histogram
plot, shown in Figure \ref{fig:histogram} where we plot each trend on
its own line but stack up the segments of the trend, much like a
histogram.

@@

The second visualization is the trend-time line plot (Figure
\ref{fig:trendtimeline}), it displays repeating trends more clearly by
dedicating a horizontal line for trend segments belonging to one
trend. This means that if a trend contains discontinuous segments then
the segments appear on the same line.  However, the least common
trends need to be pruned away or the view will be very long.

The third visualization is the histogram plot (Figure
\ref{fig:histogram}) which superficially resembles the time-trend
plot.  However, in this view the trends are plotted together by
stacking to the left of their row, thus time information is lost.  The
trends are ordered by the number of topics in the trend.  The
histogram plot shows the count of instances of a trend and thus
indicates which trends occur the most. Unfortunately due to the large
number of topics given allotted space it is often best to crop off the
trends with only one topic, otherwise the tail is long.

%XXX Describe something useful like a trend
\begin{figure*}
  \centering
  \includegraphics[width=1.0\textwidth]{fixed-time-smear-plot-scaled}
  \caption{Compact trend plot shows topics per month for MaxDB
    7.500. The x-axis is time in months, the Y axis is used to stack
    topics occurring at the same time. Trends that are continuous are
    plotted as continuous blocks. The top 10 words in the topics are
    joined and embedded in box representing the topics.}
  \label{fig:topicsmear}
\end{figure*}

\begin{figure}
  \centering
  \includegraphics[width=0.45\textwidth]{fixed-time-smear-plot-cropped}
  \caption{A zoomed in slice of compact trend plot of topics per month of MaxDB 7.500. The topic text is visible in each topic box. Trends are plotted across time continuously.}
  \label{fig:zoomedsmear}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.45\textwidth]{class-smear-plot-crop-scaled}
  \caption{Trend-time Line: Trends plotted per month of MaxDB 7.500. Time in months are plotted along the x-Axis, each row on the y-axis is associated with a trend ranked by size in descending order}         
  \label{fig:trendtimeline}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.45\textwidth]{histogram-cropped-scaled}
  \caption{The top part of a trend histogram of MaxDB 7.500, ordered
    by topic occurrence. X axis determines the number of continuous
    months of a trend. Trends are ranked by the number of topics that
    a trend contains in descending order.}
  \label{fig:histogram}
\end{figure}

@@

When we examined PostgreSQL, we did not find many trends with two or
more topics with a similarity of $7/10$. These trends were not very
large, lasting on 3 months at most. The first, second largest trends
were dedicated to the work of two external developers: Dal Zotto, Dan
McGuirk. The fifth largest trend related to work by PostgreSQL developer D'Arcy. Other
topics of the larger trends were changes to the TODO list in HTML, and
time string formating topics relating to timezones.

If we kept the stop words we found that the large trend consisted
mostly of stop words and non-stop words such as \emph{patch},
\emph{fix}, \emph{update}.  By decreasing the similarity constraint to
$1/2$ similarity the largest most the largest most common trend, which
stretched across development, contained these same words
(\emph{patch}, \emph{fix}, \emph{update}). The second most common
topic mentions Dal Zotto, while the third most common topic mentions
the \texttt{[PATCHES]} mailinglist and the names of some patch
contributors.  Other repeating topics refer to portability with Win32,
Central Europe Time (CEST) from email headers, issues with ALTER TABLE
and CVS branch merging (because CVS doesn't record merges explicitly).

% When we examined PostgreSQL, we did not find many trends with two or
% more topics; however, we did find one large trend which stretched
% across most of its existence.  Much of the large trend consisted of
% stop words but it did include topics which had non-stop words such as
% \emph{patch}, \emph{fix}, \emph{update}.

% The second largest topic was interesting because it concerned changes from
% 1998, from developers using Central Europe Time (CEST), where the two
% notable tokens were 1998 and CEST. These changes appear to be patches
% submitted by one particular developer who lived in that timezone.

% The third largest trend we found appeared to concerned with copying
% and moving files between branches as it referenced the old branch.
% branches.  We inferred this trend based on the fact that the old
% version identifier was referenced, and we know this this is common
% practice when performing a branch merge, since CVS cannot record such
% information directly.

% Once stop words were removed the some of the purposes became more
% clear. The large trend disappeared, the largest trend became bug fixes
% submitted by a developer named Dal Zotto which is a short trend.

@@

We tracked Firebird from August 2000 to January 2006. We found that with a
similarity of $7/10$ that Firebird had far more continuous and
recurring trends than PosgreSQL.  The first large trend was segmented
across time but explicitly references one author \texttt{carlosga05}
and words like \emph{added}, \emph{fixed} and \emph{updated}.

The second largest trend was during the month of March 2001 relating
to incremental building and the porting of developer Konstantin's
Solaris port of the Firebird build files. The third largest trend was
about JDBC, which is how Firebird and Java communicate.  Other trends
included topics regarding AIX PPC compilation, updating the build
process, internationalization and UTF8, Darwin build support and bug
fixing.

Topics that were not trends but appeared to be interesting were mostly
external bug fixes submitted to the project.  In these cases, the
developers would express gratitude in their commit log comments, such
as ``Thanks, Bill Lam''.  Other easily discernible topics included
tokens and phrases such as: \emph{compiler workarounds}, \emph{nightly
  updates}, \emph{packets} and \emph{MSVC++}.

% We tracked Firebird from August 2000 to January 2006.
% Our first attempt with Firebird resulted in a set of trends that were
% dominated by one large trend containing tokens that were mostly stop
% words such as references to one author
% \texttt{carlosga05} and the words: \emph{the, to, a, for, of, is, it}.

% The second largest trend related to branching and possibly merging; it
% contained language such as ``was added on branch'' and contained
% references to tags and branches.

% The third largest trend contained references to the changelog, the
% build system, architecture and platform support. It specifically
% mentioned tokens such as \emph{Solaris}, \emph{port} and \emph{x86}.

% Other trends included topics regarding AIX PPC compilation, updating
% the build process, internationalization and UTF8, Darwin build support
% and bug fixing.

% Topics that were not trends but appeared to be  interesting were mostly external bug
% fixes submitted to the project.
%  In these cases, the developers would express thanks in their commit log comments, such as ``Thanks, Bill Lam''.  Other easily discernible topics included 
%  tokens and topics such as: \emph{compiler workarounds}, \emph{nightly updates}, \emph{packets} and \emph{MSVC++}.

@@

The plots we produced of MaxDB 7.500 were unlike those of the other
systems, as there was a period where no real development occurred and
thus there were no topics or trends whatsoever. Using a topic
similarity of $7/10$ we evaluated of MaxDB 7.500, the first period
from June 2004 to January 2005, and the second period from June 2005
to June 2006.

The largest common trend has references to build system files like
\texttt{SYSDD.punix} and \texttt{MONITOR.punix}.  This trend is
partially displayed at the top of the zoomed in compressed view plot
(Figure \ref{fig:zoomedsmear}) and at the top of the histogram plot
(Figure \ref{fig:histogram}).  Other tokens mentioned are
\emph{Sutcheck v1.09}, the prefix SUT stands for Storable Unit
Type. Sutcheck would also automate check-ins using a Perforce SCS tool
which was exporting check-ins to CVS.

@@

The third most common trend, seen on Figure \ref{fig:trendtimeline},
seemed to include tokens related to operating system support, such as
Linux and Windows, as well as architecture support, AMD64 and
Opteron. The word \emph{problem} was common among all of these
trends. This trend seemed related to the smaller fourth largest trend
that had tokens \emph{AMD64} and \emph{Windows}. This shows that
topics can overlap but still not match via our similarity measure.

%XXX Edit
% Was reading lda.report
Bug tracker URLs dominated unique topics during some months. For
instance in the last month of MaxDB 7.500 development every topic
contained 1 unique Bug tracker URL, this pattern did not occur in the
previous month. We investigated the actual revisions and we did find
there was difference in behavior between and that the last month of
development commits were mostly referencing the bug tracker. If the
topics of one month were 20 or more unique bug tickets being
addressed, the global LDA analysis would probably miss this, yet these
bug tickets were the focus of development for that month and not
necessarily globally relevant.

@@

We noticed that commits which mentioned \emph{PHP} occured two thirds
less frequently than commits which mentioned \emph{Perl}, but
\emph{Perl} related topics appeared in the global static topics for
MaxDB while \emph{PHP} related topics did not.  Our local topic
analysis mentioned \emph{PHP} in 5 different topics, yet only
mentioned \emph{Perl} in 4 different topics. Perhaps the Perl mentions
were not as independent topics as the PHP topics were.

We noticed that Perforce and Changelist were in just about every
topic, so we added them to the stop words list. Longer trends were
shortened and in some cases the total number of topics found per month
was less. Evaluating different similarity thresholds showed that by
removing common words one reduces the general similarity of topics,
that said the larger topics still existed. Thus if you add more
relevant stop words one should tune your topic similarity parameters
to handle these changes.

@@

If we approach the difference of static global topic analysis and
windowed topic analysis via common tokens we can see that common token
tend to dominate both results, at least in the case of MaxDB 7.500 our
local topic approach netted us topics which contained these relatively
important and common words: \emph{UNICODE}, \emph{DEC/OSF1},
\emph{ODBC}, \emph{crash}, \emph{SQLCLI}, \emph{SYSDD.cpnix},
\emph{backup}, \emph{select}, \emph{make}, \emph{memory}, \emph{view},
and finally \emph{debug}. None of these tokens were part of the static
topic analysis topics, but they were part of 566 commits (6\% of the
entire system) to MaxDB 7.500. These tokens were part of 87 out of 500
(25 months, 20 topics per month) of our locally generated topics.

@@

Each project seemed to have their own kind of stop words, we did not
remove them, but perhaps dropping some of these stop words would aid
the clarity of topics produced through topic analysis; alternatively
it might cause our topic similarity plots to be fundamentally
different.

@@

We proposed and demonstrated the application of windowed topic
analysis, that is, topic analysis applied to commit messages during
periods of development. This approach differed from previous work that
would apply topic analysis globally to a project. We showed that many
topics that exist locally are relevant and interesting yet would often
not be included in global topic analysis. We identify recurring topics
with a topic similarity measure which looks for topics which recur and
mutate throughout the development of the software project.

The value of windowed topic analysis was demonstrated on a case study
of three database systems with a particular focus on MaxDB 7.500. We
showed that static topic analysis missed important topics like ODBC
changes.

@@

We presented multiple visualizations that focused on different aspects
of the data: temporality of trends, trend size, and a compressed
overall trend view. The compressed trend view shows much more
information than the views that global analysis could show, it
indicates how focused a period is by the total number of topics as
well shows topics by similarity so one can track trends across time.
Our histogram trend-view immediately hi-lights and measures the size of
trends while our time-line trend view shows how a topic reoccurs over
time. These visualizations help us get an over feeling about the
common and unique topics that developers focus on. If implemented
properly a user can easily zoom and query a topic.
